Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-12 manuscript/main.tex
orthonormal proposition intro revised

Modern biomedical studies frequently collect diverse data types from each subject.
As an illustrative example, the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021}  records both multiple renogram curves (functional data) and multiple renogram variables (vector data)  for each kidney. These modalities provide distinct yet related physiological signals. To effectively analyze such heterogeneous data, we need statistical methods capable of jointly modeling functional and vector-valued covariates. This approach accounts for both shared and modality-specific variations linked to clinical outcomes. Our work focuses on developing a method to fit the following hybrid functional linear regression model:
\begin{equation}	\label{eq: functional model}
	Y_i
	= 
	\boldsymbol{\beta}^\top \mathbf{Z}_i
	+ 
	\sum 
	\limits_{j=1}^K 
	\int  \beta_j(t) X_{ij}(t) dt + \epsilon_i,
	\quad
	i = 1, \ldots, n,
	%= \langle \boldsymbol{\alpha}, \mathbf{Z} \rangle +  \langle \beta, X \rangle_{\mathcal{F}}  + \epsilon
\end{equation}
where $Y$ is a scalar response, $\mathbf{Z}$ is the Euclidean vector covariate and 
$X_{i1}(t), \ldots, X_{iK}(t)$ are functional predicters.
For notational convenience, we assume throughout this paper that the responses and predictors have been centered, allowing the intercept term to be ignored.
%where $\boldsymbol{\alpha}=(\alpha_1, \ldots, \alpha_p)^\top \in \mathbb{R}^p$ and $\beta=(\beta^{(1)}, \ldots, \beta^{(K)}) \in \mathcal{F}$ are respectively the regression coefficient vector and function that characterize the effect of scalar and functional predictors on the outcome.

@@

\subsection{Geometric Properties}\label{section:sub:prop_geom}\jongmin{currently revising}
In this section, we derive several geometric properties of the PLS components and and scores obtained from the proposed algorithm.

\begin{definition}[Roughness-sensitive inner product]
		Given two hybrid predictors 	$W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$
	and
	$W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$ belonging to $\mathcal{H}$ defined in Definition \ref{def:hybid_predictor}
	and roughness penalty matrix 	$\Lambda  = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$,
	The inner product between two hybrid predictors, taking into the roughness peanlties,  is then defined as:
\begin{equation}
	\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}
\end{equation}
	\begin{equation}		\label{eq: hybrid inner product}
	\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
	:=
	\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
	\sum \limits_{k=1}^K \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
	+ 
	\omega
	\mathbf{Z}_1^\top \mathbf{Z}_2,
\end{equation}
for $\omega > 0$.
The corresponding norm $\Vert \cdot \Vert_\mathcal{H}$ is derived from this inner product as $\langle \cdot,\cdot \rangle_\mathcal{H}^{1/2}$.
\end{definition}

  ($\widehat{\xi}_1, \widehat{\xi}_2, \ldots$)  based on the modified inner product  which now takes into account the roughness of the functions:
\begin{align*}
	\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
	&= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right \rangle_{L^2} \\
	&= \sum \limits_{k=1}^K \int_0^1 \widehat{\psi}_l^{(k)}(t_k) \widehat{\psi}_j^{(k)}(t_k) dt_k + \sum \limits_{r=1}^p \widehat{\theta}_{lr} \widehat{\theta}_{jr} + \sum_{k=1}^K \lambda_k \int_0^1 \ddot{\widehat{\psi}}_l^{(k)}(t_k) \ddot{\widehat{\psi}}_j^{(k)}(t_k) dt_k \\
	& = \mathbb{1}(l=j). 
\end{align*}
This modified orthonormality condition is one of the main geometric properties of the proposed PLS algorithm and is formally stated with proof in Section \ref{section:main:properties_of_our_method} (see Proposition \ref{proposition: modified orthnormality of PLS components}), along with other properties.

The following proposition states that the PLS components estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal with respect to the modified inner product $\langle \langle \cdot \rangle \rangle$ that incorporates the roughness of the functions.
\begin{proposition}\label{proposition: modified orthnormality of PLS components}
	The estimated PLS components, $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, from the proposed PLS algorithm are mutually orthonormal with respect to the modified inner product $\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}$ in $\mathcal{H}$ that incorporates the roughness of the functions; i.e.,
	\begin{equation*}
		\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
		= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right\rangle_{L^2}
		= \mathbb{1}(l=j). 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.
Now let $\widehat{\boldsymbol{\rho}}_l=(\widehat{\rho}_{1l}, \ldots, \widehat{\rho}_{nl})^\top$ denote the $n$-dimensional vector whose elements consist of estimated $l$-th PLS scores ($l=1, \ldots, L)$ of $n$ observations. The next proposition states that the vectors of estimated PLS scores $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$ obtained from the proposed algorithm are mutually orthogonal.
\begin{proposition}	\label{proposition: orthnormality of PLS scores}
	The vectors of estimated PLS scores, $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$, obtained from the proposed PLS algorithm are mutually orthogonal in the sense that
	\begin{equation*}
		\widehat{\boldsymbol{\rho}}_l^\top \widehat{\boldsymbol{\rho}}_j= 0 \quad \textnormal{for} \quad \; l,j \in \{1, \ldots, L\}, \; l \ne j. 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.

>>>>

Modern biomedical studies frequently collect diverse data types from each subject.
As an illustrative example, the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021}  records both multiple renogram curves (functional data) and multiple renogram variables (vector data)  for each kidney. These modalities provide distinct yet related physiological signals. To effectively analyze such heterogeneous data, we need statistical methods capable of jointly modeling functional and vector-valued covariates. This approach accounts for both shared and modality-specific variations linked to clinical outcomes. Our work focuses on developing a method to fit the following hybrid functional linear regression model:
\begin{equation}	\label{eq: functional model}
	Y_i
	= 
	\boldsymbol{\beta}^\top \mathbf{Z}_i
	+ 
	\sum 
	\limits_{k=1}^K 
	\int  \beta_j(t) X_{ik}(t) dt + \epsilon_i,
	\quad
	i = 1, \ldots, n,
	%= \langle \boldsymbol{\alpha}, \mathbf{Z} \rangle +  \langle \beta, X \rangle_{\mathcal{F}}  + \epsilon
\end{equation}
where $Y_i$ is a scalar response, $\mathbf{Z}_i$ is a Euclidean vector covariate, 
$X_{i1}(t), \ldots, X_{iK}(t)$ are functional predicters, and $\epsilon_i$ is an observational noise.
For notational convenience, we assume throughout this paper that the responses and predictors have been centered, allowing the intercept term to be ignored.
%where $\boldsymbol{\alpha}=(\alpha_1, \ldots, \alpha_p)^\top \in \mathbb{R}^p$ and $\beta=(\beta^{(1)}, \ldots, \beta^{(K)}) \in \mathcal{F}$ are respectively the regression coefficient vector and function that characterize the effect of scalar and functional predictors on the outcome.

@@

\subsection{Geometric Properties}\label{section:sub:prop_geom}
In this section, we derive several geometric properties of the PLS components and and scores obtained from the proposed algorithm.

A fundamental property of partial least squares   is that its derived bases are orthonormal. Our regularized estimates preserve this property, with respect to a modified inner product that incorporates the roughness penalty, defines as follows:
\begin{definition}[Roughness-sensitive inner product]\label{def:hybrid_inner_product_roughness}
	Given two hybrid predictors $W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$ and $W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, both elements of $\mathcal{H}$ as defined in Definition \ref{def:hybid_predictor}, and a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, the roughness-sensitive inner product between $W_1$ and $W_2$ is defined as:
	\begin{equation}		\label{eq: hybrid inner product_roughness}
		\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
		:=
		\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
		+
		\sum \limits_{k=1}^K \lambda_k \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
		+
		\mathbf{Z}_1^\top \mathbf{Z}_2.
	\end{equation}
\end{definition}
Based on this inner product, the following proposition states that the PLS component directions estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal.
\begin{proposition}\label{proposition: modified orthnormality of PLS components}\jongmin{currently revising}
	The estimated PLS components, $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, from the proposed PLS algorithm are mutually orthonormal with respect to the modified inner product $\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}$ in $\mathcal{H}$ that incorporates the roughness of the functions; i.e.,
	\begin{equation*}
		\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
		= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right\rangle_{L^2}
		= \mathbb{1}(l=j). 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.
Now let $\widehat{\boldsymbol{\rho}}_l=(\widehat{\rho}_{1l}, \ldots, \widehat{\rho}_{nl})^\top$ denote the $n$-dimensional vector whose elements consist of estimated $l$-th PLS scores ($l=1, \ldots, L)$ of $n$ observations.

($\widehat{\xi}_1, \widehat{\xi}_2, \ldots$)  based on the modified inner product  which now takes into account the roughness of the functions:
\begin{align*}
	\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
	&= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right \rangle_{L^2} \\
	&= \sum \limits_{k=1}^K \int_0^1 \widehat{\psi}_l^{(k)}(t_k) \widehat{\psi}_j^{(k)}(t_k) dt_k + \sum \limits_{r=1}^p \widehat{\theta}_{lr} \widehat{\theta}_{jr} + \sum_{k=1}^K \lambda_k \int_0^1 \ddot{\widehat{\psi}}_l^{(k)}(t_k) \ddot{\widehat{\psi}}_j^{(k)}(t_k) dt_k \\
	& = \mathbb{1}(l=j). 
\end{align*}

The next proposition states that the vectors of estimated PLS scores $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$ obtained from the proposed algorithm are mutually orthogonal.
\begin{proposition}	\label{proposition: orthnormality of PLS scores}
	The vectors of estimated PLS scores, $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$, obtained from the proposed PLS algorithm are mutually orthogonal in the sense that
	\begin{equation*}
		\widehat{\boldsymbol{\rho}}_l^\top \widehat{\boldsymbol{\rho}}_j= 0 \quad \textnormal{for} \quad \; l,j \in \{1, \ldots, L\}, \; l \ne j. 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.
